---
title: "A web-scraping exercise from Homework 3"
author: "Brent"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
  pdf_document:
    toc: yes
---

```{r}
#| label: load-libraries
#| echo: false # This option disables the printing of code (only output is displayed).
#| message: false
#| warning: false

library(tidyverse)
library(wbstats)
library(tictoc)
library(skimr)
library(countrycode)
library(here)
library(DBI)
library(dbplyr)
library(arrow)
library(rvest)
library(robotstxt) # check if we're allowed to scrape the data
library(scales)
library(sf)
library(readxl)
```

# Money in US politics

In the United States, [*"only American citizens (and immigrants with green cards) can contribute to federal politics, but the American divisions of foreign companies can form political action committees (PACs) and collect contributions from their American employees."*](https://www.opensecrets.org/political-action-committees-pacs/foreign-connected-pacs)

We will scrape and work with data foreign connected PACs that donate to US political campaigns. The data for foreign connected PAC contributions in the 2022 election cycle can be found at https://www.opensecrets.org/political-action-committees-pacs/foreign-connected-pacs/2022. Then, we will use a similar approach to get data such contributions from previous years so that we can examine trends over time.

All data come from [OpenSecrets.org](https://www.opensecrets.org), a *"website tracking the influence of money on U.S. politics, and how that money affects policy and citizens' lives"*.

```{r, eval=FALSE}
#| label: allow-scraping-opensecrets
#| warning: false
#| message: false

library(robotstxt)
paths_allowed("https://www.opensecrets.org")

base_url <- "https://www.opensecrets.org/political-action-committees-pacs/foreign-connected-pacs/2022"

contributions_tables <- base_url %>%
  read_html() 

contributions <- contributions_tables %>% 
  html_element(".DataTable-Partial") %>% ## select table element
  html_table()

class(contributions)

```

- First, make sure you can scrape the data for 2022. Use janitor::clean_names() to rename variables scraped using `snake_case` naming. 

```

- Clean the data: 

    -   Write a function that converts contribution amounts in `total`, `dems`, and `repubs` from character strings to numeric values.
    -   Separate the `country_of_origin_parent_company` into two such that country and parent company appear in different columns for country-level analysis.

```{r, eval=FALSE}
# write a function to parse_currency
parse_currency <- function(x){
  x %>%
    
    # remove dollar signs
    str_remove("\\$") %>%
    
    # remove all occurrences of commas
    str_remove_all(",") %>%
    
    # convert to numeric
    as.numeric()
}

# clean country/parent co and contributions 
contributions <- contributions %>%
  janitor::clean_names() %>% 
  separate(country_of_origin_parent_company, 
           into = c("country", "parent"), 
           sep = "/", 
           extra = "merge") %>%
  mutate(
    total = parse_currency(total),
    dems = parse_currency(dems),
    repubs = parse_currency(repubs)
  )

print(contributions)
```

-   Write a function called `scrape_pac()` that scrapes information from the Open Secrets webpage for foreign-connected PAC contributions in a given year. This function should

    -   have one input: the URL of the webpage and should return a data frame.
    -   add a new column to the data frame for `year`. We will want this information when we ultimately have data from all years, so this is a good time to keep track of it. Our function doesn't take a year argument, but the year is embedded in the URL, so we can extract it out of there, and add it as a new column. Use the `str_sub()` function to extract the last 4 characters from the URL. You will probably want to look at the help for this function to figure out how to specify "last 4 characters".
    
```{r}
scrape_pac<-function(base_url){
  # Get the year in the URL
  year<-str_sub(base_url,-4)
  # Read the html file in the URL
  webpage<-read_html(base_url)
  # Turn the html file in the URL into a dataframe
  data<-html_table(webpage)
  
  data_clean<- data[[1]] %>% 
    # Clean the column titles using snake case
    janitor::clean_names(case="snake") %>% 
    # Insert a year column using the year in the URL
    mutate(year=year)
  
  return(data_clean)
}

```


-   Define the URLs for 2022, 2020, and 2000 contributions. Then, test your function using these URLs as inputs. Does the function seem to do what you expected it to do?

```{r}
# URLs

url_2022<-"https://www.opensecrets.org/political-action-committees-pacs/foreign-connected-pacs/2022"
url_2020<-"https://www.opensecrets.org/political-action-committees-pacs/foreign-connected-pacs/2020"
url_2000<-"https://www.opensecrets.org/political-action-committees-pacs/foreign-connected-pacs/2000"

# Testing urls as inputs

result_2022 <- scrape_pac(url_2022)
result_2020 <- scrape_pac(url_2020)
result_2000 <- scrape_pac(url_2000)

# View the results
head(result_2022)
head(result_2020)
head(result_2000)

```


-   Construct a vector called `urls` that contains the URLs for each webpage that contains information on foreign-connected PAC contributions for a given year.

```{r}
years<-c(2000,2020,2022)

urls<- str_c("https://www.opensecrets.org/political-action-committees-pacs/foreign-connected-pacs/",years)
```
-   Map the `scrape_pac()` function over `urls` in a way that will result in a data frame called `contributions_all`.
```{r}
library(purrr)

contributions_all <- map_df(urls, scrape_pac)

print(contributions_all)
```


# Scraping consulting jobs

The website [https://www.consultancy.uk/jobs/](https://www.consultancy.uk/jobs) lists job openings for consulting jobs.

```{r}
#| label: consulting_jobs_url
#| eval: false

library(robotstxt)
paths_allowed("https://www.consultancy.uk") #is it ok to scrape?

base_url <- "https://www.consultancy.uk/jobs/page/1"

listings_html <- base_url %>%
  read_html()

```
Identify the CSS selectors in order to extract the relevant information from this page, namely

1. job 
1. firm
1. functional area
1. type

Can you get all pages of ads, and not just the first one, `https://www.consultancy.uk/jobs/page/1` into a dataframe?
```{r}
library(robotstxt)
paths_allowed("https://www.consultancy.uk") #is it ok to scrape?


listings_tbl <- listings_html %>% 
  html_element(".dataTable") %>% ## select table element
  html_table()


cons_url <- "https://www.consultancy.uk/jobs/page/"
htmls <- list()
i <- 1  # Initial value
error <- FALSE

while (!error) {
  # Check for an error condition
  print(i)
  tryCatch({
      # Code to be executed if no error condition
    url <- paste0(cons_url,i)
    print(url)
    listings_html <- 
      paste0(cons_url,i) %>%
      read_html()
    
    htmls <<- c(htmls, listings_html)

  }, error = function(err) {
     print("Error") 
     error <<- TRUE
  })
    
  # Update the loop variable
  i <- i + 1
}
```


-   Write a function called `scrape_jobs()` that scrapes information from the webpage for consulting positions. This function should

    -   have one input: the URL of the webpage and should return a data frame with four columns (variables): job, firm, functional area, and type

    -   Test your function works with other pages too, e.g., https://www.consultancy.uk/jobs/page/2. Does the function seem to do what you expected it to do?

    -   Given that you have to scrape `...jobs/page/1`, `...jobs/page/2`, etc., define your URL so you can join multiple stings into one string, using `str_c()`. For instnace, if `page` is 5, what do you expect the following code to produce?
    
```
base_url <- "https://www.consultancy.uk/jobs/page/1"
url <- str_c(base_url, page)
```
-   Construct a vector called `pages` that contains the numbers for each page available


-   Map the `scrape_jobs()` function over `pages` in a way that will result in a data frame called `all_consulting_jobs`.

-   Write the data frame to a csv file called `all_consulting_jobs.csv` in the `data` folder.
```{r}
# write a function to parse_currency

scrape_job <- function(url){
  print(url)
  listings_html <- 
    url %>%
    read_html()
  
  listings_tbl <- listings_html %>% 
    html_element(".dataTable") %>% ## select table element
    html_table()

  # clean country/parent co and contributions 
  listings_tbl <- listings_tbl %>%
    janitor::clean_names()
  listings_tbl
}

scrape_jobs <- function(base_url){
  error <- FALSE
  df_list <- list()
  i <- 1  # Initial value

  while (!error) {
  # Check for an error condition
  print(i)
  tryCatch({
    # Code to be executed if no error condition
    cons_url <- paste0(base_url,i)
    listings_tbl <- scrape_job(cons_url)
    
    df_list <- append(df_list, list(listings_tbl))


  }, error = function(err) {
     message("Error:", conditionMessage(err))
     print("Error") 
     error <<- TRUE
  })
    # Update the loop variable
    i <- i + 1
  }

  final_df <- bind_rows(df_list)
  print(final_df)
}
  
all_jobs <- scrape_jobs("https://www.consultancy.uk/jobs/page/")
  
print(all_jobs %>%  head(5))

```



